Using Metric Space Indexing for Complete and Efficient Record Linkage

### 2017-11-21 paper.tex
Update on Overleaf.

% TODO:
% - clearly point out novelty of this work
% - contribution compared to earlier work

@@

\textbf{M-tree:} The M-tree linkage algorithm has no additional
parameters. In a similar manner to the \emph{LSH-MinHash} approach, each
record in A is inserted into an M-tree. To perform linkage, the
algorithm iterates over each record $b$ in $B$. A $range-search(b,d)$
operation is performed, passing the distance threshold as the second
parameter. All the returned records are directly recorded as links.

@@

\begin{table}[t]
\caption{Characteristics of datasets used in the experiments.}
 \label{table-datasets}
  \centering
  \begin{scriptsize}
  %\addtolength{\tabcolsep}{-0.5pt}
  \begin{tabular}{ccccc}
  \hline\noalign{\smallskip}
  Dataset~ & ~Records in~& ~Records in~ & ~Number of true~& ~Entities\\
  name(s)  & dataset1  & dataset2  & matching pairs & linked \\
  \noalign{\smallskip} \hline \noalign{\smallskip}
  CORA  & 1,295 & 1295 & 17,184 & publication-publication\\
  Isle of Skye & 17,612 & 12,284& 2,900 & birth-death\\
  Kilmarnock  & 38,430 & 23,714 & 8,300 & birth-death\\
  \noalign{\smallskip} \hline
  \end{tabular}
  \end{scriptsize}
\end{table}

@@

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

The parameter settings used in our experiments are summarized in
Table~\ref{table-parameters}.

In all of our experiments we use a single distance metric which is the sum of the Levenshtein distances between the set of fields being compared.

@@

In the traditional blocking experiment the following fields from the Cora dataset are used as a blocking key: \textit{author}, \textit{title}, \textit{venue}, \textit{location}, \textit{publisher} and \textit{year}. We also use a very permissive blocking key which takes the union of all the values for these fields.

\Cref{comparison-of-results-cora} shows the results of the Cora experiments. The table shows the parameters for the experiment, the number of distance comparisons made, the precision, recall and F1 measure and the \textit{pairs quality} and\textit{ pairs completeness} achieved by each algorithm.  The linker column may be interpreted as follows: the algorithm name is followed by its parameters, for LSH this is the number of the bands followed by the size of bands. For traditional blocking the parameters are the fields used for blocking. \textit{Pairs quality} and \textit{pairs completeness}  are given to illustrate the efficacy of the algorithms and are defined as follows: XXXXXXXXXXX.

\textit{Pairs quality}  relates to recall and gives an indication of the number of (correct) matches amongst the links. 
Pairs completeness BLAH BLAH.

@@

TODO: Explain pairs-quality and pairs-completeness.
\emph{\textbf{TODO: discuss measures used. num comparisons, time? etc.}}

\begin{table}[ht]
\centering
\begin{tabular}{l|r|r|r|r|r|r}
\parbox{0.18\linewidth}{\centering Linker} &
\parbox{0.16\linewidth}{\centering Comparisons} &
\parbox{0.11\linewidth}{\centering Pairs Quality} &
\parbox{0.17\linewidth}{\centering Pairs \\ Completeness} &
\parbox{0.11\linewidth}{\centering Precision} &
\parbox{0.08\linewidth}{\centering Recall} &
\parbox{0.11\linewidth}{\centering F1 Measure} \\ \hline \hline
Brute Force        & 1,677,025                    & 0.02                  & 0.33                       & 0.84      & 0.57   & 0.68       \\ \hline
MTree              &  902,693                     & 0.89                  & 0.22                       & 0.84      & 0.57   & 0.68       \\ \hline
LSH-2-2            &  175,435                     & 0.31                  & 0.29                       & 0.94      & 0.53   & 0.68       \\
LSH-5-2            &  175,435                     & 0.31                  & 0.29                       & 0.94      & 0.53   & 0.68       \\
LSH-10-2           &  175,435                     & 0.31                  & 0.29                       & 0.94      & 0.53   & 0.68       \\
LSH-2-5            &  175,435                     & 0.31                  & 0.29                       & 0.94      & 0.53   & 0.68       \\
LSH-5-5            &  175,435                     & 0.31                  & 0.29                       & 0.94      & 0.53   & 0.68       \\
LSH-10-5           &  175,435                     & 0.31                  & 0.29                       & 0.94      & 0.53   & 0.68       \\
LSH-2-10           &  175,435                     & 0.31                  & 0.29                       & 0.94      & 0.53   & 0.68       \\
LSH-5-10           &  175,435                     & 0.31                  & 0.29                       & 0.94      & 0.53   & 0.68       \\
LSH-10-10          &  175,435                     & 0.31                  & 0.29                       & 0.94      & 0.53   & 0.68       \\  \hline
Block-year      &  115,893                     & 0.22                  & 0.19                       & 0.99      & 0.35   & 0.51       \\
Block-authors   &   11,039                      & 0.74                  & 0.1                        & 0.94      & 0.16   & 0.28       \\
Block-title     &   27,407                      & 0.84                  & 0.23                       & 0.95      & 0.42   & 0.58       \\
Block-venue     &   36,647                      & 0.7                   & 0.12                       & 0.85      & 0.29   & 0.44       \\
Block-location  & 1,009,957                    & 0.17                  & 0.24                       & 0.83      & 0.43   & 0.57       \\
Block-publisher &  833,079                     & 0.12                  & 0.23                       & 0.85      & 0.44   & 0.58       \\
Block-combined  & 1,214,269                    & 0.05                  & 0.32                       & 0.84      & 0.56   & 0.67       \\ \hline
\end{tabular}
\caption{Linkage Quality on the Cora dataset distance=70\label{comparison-of-results-cora}
\textbf{TODO: The table needs some rows removed. Note that this table is with a threshold value of 70.}
}
\end{table}

@@

\begin{table}[ht]
\centering
\begin{tabular}{l|r|r|r|r|r|r}
\parbox{0.18\linewidth}{\centering Linker} &
\parbox{0.16\linewidth}{\centering Comparisons} &
\parbox{0.11\linewidth}{\centering Pairs Quality} &
\parbox{0.17\linewidth}{\centering Pairs \\ Completeness} &
\parbox{0.11\linewidth}{\centering Precision} &
\parbox{0.08\linewidth}{\centering Recall} &
\parbox{0.11\linewidth}{\centering F1 Measure} \\ \hline \hline
MTree    & 102,318,525         & 0.84                  & 0.86                       & 0.65      & 0.46   & 0.54      \\ \hline
LSH-2-2  & 3,109,250           & 0                     & 0.05                       & 0.63      & 0.03   & 0.06       \\
LSH-5-2  & 10,412,496          & 0                     & 0.16                       & 0.64      & 0.11   & 0.19       \\
LSH-10-2 & 53,874,127          & 0                     & 0.46                       & 0.68      & 0.36   & 0.47       \\
LSH-5-5  & 36,566             & 0                     & 0.01                       & 0.76      & 0.01   & 0.01       \\
LSH-10-5 & 129,873            & 0                     & 0.01                       & 0.72      & 0.01   & 0.02       \\ \hline
\end{tabular}
\caption{Linkage Quality on demography datasets \textbf{This is for Skye. Threshold value is 2.}
\label{comparison-of-results-demography-skye}
}
\end{table}

\begin{table}[ht]
\centering
\begin{tabular}{l|r|r|r|r|r|r}
\parbox{0.18\linewidth}{\centering Linker} &
\parbox{0.16\linewidth}{\centering Comparisons} &
\parbox{0.11\linewidth}{\centering Pairs Quality} &
\parbox{0.17\linewidth}{\centering Pairs \\ Completeness} &
\parbox{0.11\linewidth}{\centering Precision} &
\parbox{0.08\linewidth}{\centering Recall} &
\parbox{0.11\linewidth}{\centering F1 Measure} \\ \hline \hline
MTree     & 514,871,153         & 0.89                  & 0.82                       & 0.76      & 0.45   & 0.57      \\ \hline
LSH-2-2   & 99,145,887          & 0                     & 0.23                       & 0.81      & 0.16   & 0.27       \\
LSH-5-2   & 130,721,338         & 0                     & 0.32                       & 0.79      & 0.23   & 0.36       \\
LSH-10-2  & 177,168,848         & 0                     & 0.48                       & 0.79      & 0.36   & 0.49       \\
LSH-5-5   & 239,368            & 0.01                  & 0.02                       & 0.84      & 0.01   & 0.02       \\
LSH-10-5  & 855,431            & 0.01                  & 0.03                       & 0.87      & 0.02   & 0.03 \\ \hline
\end{tabular}
\caption{Linkage Quality on demography datasets\textbf{This is for Kilmarnock. Threshold value is 2.}
\label{comparison-of-results-demography-kili}
}
\end{table}

>>>>

@@

\textbf{M-tree:} The M-tree linkage algorithm has no additional
parameters. In a similar manner to the \emph{LSH-MinHash} approach, each
record in A is inserted into an M-tree. To perform linkage, the
algorithm iterates over each record $b$ in $B$. A \emph{range-search(b,d)}
operation is performed, passing the distance threshold as the second
parameter. All the returned records are directly recorded as links.

@@

\begin{table}[t]
\caption{Characteristics of datasets used in the experiments.}
 \label{table-datasets}
  \centering
  \begin{scriptsize}
  %\addtolength{\tabcolsep}{-0.5pt}
  \begin{tabular}{ccccc}
  \hline\noalign{\smallskip}
  Dataset~ & ~Records in~& ~Records in~ & ~Number of true~& ~Entities\\
  name(s)  & dataset 1  & dataset 2  & matching pairs & linked \\
  \noalign{\smallskip} \hline \noalign{\smallskip}
  Cora & 1,295 & 1,295 & 17,184 & publication-publication\\
  Isle of Skye & 17,612 & 12,284& 2,900 & birth-death\\
  Kilmarnock & 38,430 & 23,714 & 8,300 & birth-death\\
  \noalign{\smallskip} \hline
  \end{tabular}
  \end{scriptsize}
\end{table}

@@

In all of our experiments we use a single distance metric which is the
sum of the Levenshtein~\cite{Levenshtein66} distances between the set of
fields being compared.

@@

In the traditional blocking experiment the following fields from the Cora dataset are used as individually as blocking keys: \textit{author}, \textit{title}, \textit{venue}, \textit{location}, \textit{publisher} and \textit{year}. We also use a combined blocking key comprising these fields.

\Cref{comparison-of-results-cora} shows the results of the Cora experiments. The table shows the parameters for the experiment, the number of distance comparisons made, the precision, recall and the F1 measure achieved by each algorithm.  The linker column may be interpreted as follows: the algorithm name is followed by its parameters, for LSH this is the number of the bands followed by the size of bands. For traditional blocking the parameters are the fields used for blocking.

@@

\emph{\textbf{TODO: discuss measures used. num comparisons, time? etc.}}

\begin{table}[ht]
\centering
\begin{tabular}{l|r|r|r|r}
\parbox{0.2\textwidth}{\centering Linker} &
\parbox{0.2\textwidth}{\centering Comparisons} &
\parbox{0.17\textwidth}{\centering Precision} &
\parbox{0.17\textwidth}{\centering Recall} &
\parbox{0.17\textwidth}{\centering F1 Measure} \\ \hline \hline
Brute Force        & 1,677,025                 & 0.84      & 0.57   & 0.68       \\ \hline
MTree              &  902,693                  & 0.84      & 0.57   & 0.68       \\ \hline
LSH-2-2            &  192,199	& 0.95	& 0.47	& 0.63       \\
LSH-5-2            &  342,849	& 0.91	& 0.55	& 0.69       \\
LSH-10-2           &  513,947	& 0.88	& 0.57	& 0.69       \\
LSH-2-5            &  14,329	& 0.99	& 0.28	& 0.43      \\
LSH-5-5            &  22,057	& 0.99	& 0.36	& 0.53       \\
LSH-10-5           &  26,167	& 0.98	& 0.4	& 0.57      \\
LSH-2-10           &  4,711	& 1	& 0.15	& 0.27    \\
LSH-5-10           &  6,501	& 1	& 0.19	& 0.32      \\
LSH-10-10          & 10,627  	& 0.99	& 0.27& 	0.43       \\  \hline
Block-year         &  115,893                  & 0.99      & 0.35   & 0.51       \\
Block-authors      &   11,039                   & 0.94      & 0.16   & 0.28       \\
Block-title        &   27,407                   & 0.95      & 0.42   & 0.58       \\
Block-venue        &   36,647                   & 0.85      & 0.29   & 0.44       \\
Block-location     & 1,009,957                 & 0.83      & 0.43   & 0.57       \\
Block-publisher    &  833,079                  & 0.85      & 0.44   & 0.58       \\
Block-combined     & 1,214,269                 & 0.84      & 0.56   & 0.67       \\ \hline
\end{tabular}
\caption{Linkage Quality on the Cora dataset distance=70\label{comparison-of-results-cora}
\textbf{TODO: The table needs some rows removed. Note that this table is with a threshold value of 70.}
}
\end{table}

@@

\begin{table}[ht]
\centering
\begin{tabular}{l|r|r|r|r}
\parbox{0.2\textwidth}{\centering Linker} &
\parbox{0.2\textwidth}{\centering Comparisons} &
\parbox{0.17\textwidth}{\centering Precision} &
\parbox{0.17\textwidth}{\centering Recall} &
\parbox{0.17\textwidth}{\centering F1 Measure} \\ \hline \hline
MTree    & 102,318,525                 & 0.65      & 0.46   & 0.54      \\ \hline

LSH-2-2  & 3,109,250                   & 0.63      & 0.03   & 0.06       \\

LSH-5-2  & 10,412,496                  & 0.64      & 0.11   & 0.19       \\

LSH-10-2 & 53,874,127                  & 0.68      & 0.36   & 0.47       \\

LSH-5-5  & 36,566                     & 0.76      & 0.01   & 0.01       \\

LSH-10-5 & 129,873                    & 0.72      & 0.01   & 0.02       \\ \hline
\end{tabular}
\caption{Linkage Quality on demography datasets \textbf{This is for Skye. Threshold value is 2.}
\label{comparison-of-results-demography-skye}
}
\end{table}

\Cref{comparison-of-results-demography-skye} and \Cref{comparison-of-results-demography-kili} shows the results for LSH-Minhash and M-tree applied to The Isle of Skye and Kilmarnock datasets. The shingle-size was set to 2 for all the LSH experiments shown as this was found to give good results and LSH was not especially sensitive to this parameter. Results for other shingle sizes were omitted from this paper for brevity. In these experiments the linkage performed is between birth and death records. In both cases the F1 measure achieved is better for M-tree than any of the experimental settings chosen for LSH. This is largely due to the recall for M-tree being much higher than any of the LSH experiments. For all LSH parameters LSH out performs M-tree in terms of precision.

Perhaps more important than the fact that M-tree gives better F1 measures than all of the LSH experiments is the fact that the LSH performance is extremely variable. For a relatively plausible settings for the number of bands and band-size the F1 measurements vary from 0.01 (extremely poor) to 0.47 (relatively good) for Skye and between 0.03 and 0.49 for the Kilmarnock dataset. In both cases LSH-10-2 performs best, but since this is data dependent there is no guarantee these parameters would work well with another dataset. Furthermore, without good ground truth, which is usually unavailable, there is no way of knowing if good linkage results are being obtained or not.

Another observation that is worthy of comment is the number of (distance) comparisons being performed by the different algorithms. For M-trees the number of comparisons are closely related to the complexity of the search being performed since the algorithms are performing (range) searches over distance measures to determine the query results. This is not true of LSH which is performing Jaccard similarity using Hashes over the datasets. Comparisons are only used at the final step to determine if a candidate is a match or not. The LSH results which provide the best results by F1 measure are performing between a quarter and half the number of matches performed by M-trees. From this we can conclude that the hash algorithms are returning many candidates that are outwith the search distance. Thus, in order to get good results, LSH is tending towards a brute force search over the candidate results. Despite this LSH is much faster due to the efficiency of the hashing mechanisms.

\begin{table}[ht]
\centering
\begin{tabular}{l|r|r|r|r|r|r}
\parbox{0.2\textwidth}{\centering Linker} &
\parbox{0.2\textwidth}{\centering Comparisons} &
\parbox{0.17\textwidth}{\centering Precision} &
\parbox{0.17\textwidth}{\centering Recall} &
\parbox{0.17\textwidth}{\centering F1 Measure} \\ \hline \hline
MTree     & 514,871,153                   & 0.76      & 0.45   & 0.57      \\ \hline
LSH-2-2   & 99,145,887                    & 0.81      & 0.16   & 0.27       \\
LSH-5-2   & 130,721,338                   & 0.79      & 0.23   & 0.36       \\
LSH-10-2  & 177,168,848                   & 0.79      & 0.36   & 0.49       \\
LSH-5-5   & 239,368                      & 0.84      & 0.01   & 0.02       \\
LSH-10-5  & 855,431                      & 0.87      & 0.02   & 0.03 \\ \hline
\end{tabular}
\caption{Linkage Quality on demography datasets\textbf{This is for Kilmarnock. Threshold value is 2.}
\label{comparison-of-results-demography-kili}
}
\end{table}
